---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%")

library(pcal)
```

# pcal: Calibration of P-Values for Point Null Hypothesis Testing

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/pcal)](https://CRAN.R-project.org/package=pcal)
[![Build Status](https://travis-ci.org/pedro-teles-fonseca/pcal.svg?branch=master)](https://travis-ci.org/pedro-teles-fonseca/pcal) [![R build status](https://github.com/pedro-teles-fonseca/pcal/workflows/R-CMD-check/badge.svg)](https://github.com/pedro-teles-fonseca/pcal/actions)
![pkgdown](https://github.com/pedro-teles-fonseca/pcal/workflows/pkgdown/badge.svg)
[![codecov](https://codecov.io/gh/pedro-teles-fonseca/pcal/branch/master/graph/badge.svg)](https://codecov.io/gh/pedro-teles-fonseca/pcal)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

## Overview

P-values are the most commonly used tool to measure the evidence provided by the data against a model or hypothesis. Unfortunately, p-values are often incorrectly interpreted as the probability that the null hypothesis is true or as type I error probabilities. The `pcal` package uses the calibrations developed in @sellke2001 to calibrate p-values under a robust perspective and obtain measures of the evidence provided by the data in favor of point null hypotheses which are safer and more straightforward interpret: 

* `pcal()` calibrates p-values so that they can be directly interpreted as either lower bounds on the posterior probabilities of point null hypotheses or as lower bounds on type I error probabilities. With this calibration one need not fear the misinterpretation of a type I error probability as the probability that the null hypothesis is true because they coincide. Note that the output of this calibration has both Bayesian and Frequentist interpretations.
   
* `bcal()` calibrates p-values so that they can be interpreted as lower bounds on Bayes factors in favor of point null hypotheses.

Some utility functions are also included:

* `bfactor_to_prob()` turns Bayes factors into posterior probabilities using a formula from @bergerDelampady1987
  
* `bfactor_interpret()` classifies the strength of the evidence implied by a Bayes factor according to the interpretation scale suggested by @jeffreys1961
  
* `bfactor_interpret_kr()` classifies the strength of the evidence implied by a Bayes factor according to an alternative scale suggested by @kass1995
  
* `bfactor_log_interpret()` is similar to `bfactor_interpret()` but takes logarithms of Bayes factors as input

* `bfactor_log_interpret_kr()` is similar to `bfactor_interpret_kr()` but takes logarithms of Bayes factors as input

* `check_prob` checks if an object is a numeric vector of valid probability (or p-value) values. 

* `check_bf` checks if an object is a numeric vector of valid Bayes factor values. 

* `check_log_bf` checks if an object is a numeric vector of valid logarithmic Bayes factor values. It is similar to `check_log_bf` but does not require values to be non-negative. 
     
## Installation

The released version of `pcal` can be installed from [CRAN](https://CRAN.R-project.org) with:

```{r,  eval = FALSE}
install.packages("pcal")
```

The development version of `pcal` can be installed from [GitHub](https://github.com/) using the
`devtools` package:

```{r,  eval = FALSE}
# install.packages("devtools")
devtools::install_github("pedro-teles-fonseca/pcal")
```

## Usage 

First we need a p-value from any statistical test of a point null hypothesis, for example:

```{r}
x <- matrix(c(22, 13, 13, 23), ncol = 2)
pv <- chisq.test(x)[["p.value"]]

pv
```

In classical hypothesis testing, if the typical 0.05 significance threshold is used then this p-value slightly below 0.05 would result in the rejection of the null hypothesis.

With `bcal()` we can turn `pv` into a lower bound for the Bayes factor in favor of the null hypothesis:

```{r}
bcal(pv)
```

We can also turn `pv` into a lower bound for the posterior probability of the null hypothesis using `pcal()`:

```{r}
pcal(pv)
```

This is an approximation to the minimum posterior probability of the null hypothesis that we would find by changing the prior distribution of the parameter of interest (under the alternative hypothesis) over wide classes of distributions. The output of `bcal()` has an analogous interpretation for Bayes factors (instead of posterior probabilities).

Note that according to `pcal()` the posterior probability that the null hypothesis is true is at least 0.27 (approximately), which implies that a p-value below 0.05 is not necessarily indicative of strong evidence against the null hypothesis.  

One can avoid the specification of prior probabilities for the hypotheses by focusing solely on Bayes factors. To compute posterior probabilities for the hypotheses, however, prior probabilities must by specified. By default, `pcal()` assigns a prior probability of 0.5 to the null hypothesis. We can specify different prior probabilities, for example:

```{r}
pcal(pv, prior_prob = .95)
```

In this case we obtain a higher lower bound because the null hypothesis has a higher prior probability.

@sellke2001 noted that a scenario in which they definitely recommend the aforementioned calibrations is when investigating fit to the null model with no explicit alternative in mind. @pericchiTorres2011 warned that despite the usefulness and appropriateness of these p-value calibrations they do not depend on sample size, and hence the lower bounds obtained with large samples may be conservative.

Since the output of `bcal(pv)` is a Bayes factor, we can use `bfactor_to_prob()` to turn it into a posterior probability:

```{r}
bfactor_to_prob(bcal(pv)) # same as pcal(pv)

```

Like `pcal()`, `bfactor_to_prob()` assumes a prior probability of 0.5 to the null hypothesis. We can change this default:

```{r}
bfactor_to_prob(bcal(pv), prior_prob = .95)
```

To classify the strength of the evidence in favor of the null hypothesis implied by a Bayes factor we can use `bfactor_interpret()`:

```{r}
bfactor_interpret(c(0.1, 2, 5, 20, 50, 150))
```

Alternatively, we can use `bfactor_interpret_kr()`:

```{r}
bfactor_interpret_kr(c(0.1, 2, 5, 20, 50, 150))
```

Because Bayes factors are often reported on a logarithmic scale, there are also `bfactor_log_interpret()` and `bfactor_log_interpret_kr()` functions that interpret the logarithms of Bayes factors:

```{r}
bfs <- log10(c(0.1, 2, 5, 20, 50, 150))
 
bfactor_log_interpret(bfs, base = 10)

bfactor_log_interpret_kr(bfs, base = 10)
```

To compare the results with those from standard likelihood ratio tests it can be useful to obtain the strength of the evidence against the null hypothesis. If `bf` is a Bayes factor in favor  of the null hypothesis, one can use `1/bf` as input to obtain the strength of the evidence against the null hypothesis:
 
```{r}
# Evidence in favor of the null hypothesis
bfactor_interpret(c(0.1, 2, 5, 20, 50, 150))

# Evidence against the null hypothesis
bfactor_interpret(1/c(0.1, 2, 5, 20, 50, 150))
```

## Getting Help

If you find a bug, please file an issue with a minimal reproducible example on
[GitHub](https://github.com/pedro-teles-fonseca/pcal/issues). Feature requests are also welcome. You can contact me at pedro.teles.fonseca@outlook.com.

## References 




